%for a more compact document, add the option openany to avoid
%starting all chapters on odd numbered pages
\documentclass[12pt]{cmuthesis}

% This is a template for a CMU thesis.  It is 18 pages without any content :-)
% The source for this is pulled from a variety of sources and people.
% Here's a partial list of people who may or may have not contributed:
%
%        bnoble   = Brian Noble
%        caruana  = Rich Caruana
%        colohan  = Chris Colohan
%        jab      = Justin Boyan
%        josullvn = Joseph O'Sullivan
%        jrs      = Jonathan Shewchuk
%        kosak    = Corey Kosak
%        mjz      = Matt Zekauskas (mattz@cs)
%        pdinda   = Peter Dinda
%        pfr      = Patrick Riley
%        dkoes = David Koes (me)

% My main contribution is putting everything into a single class files and small
% template since I prefer this to some complicated sprawling directory tree with
% makefiles.

% some useful packages
\usepackage{times}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage[numbers,sort]{natbib}
%  \usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
%  %pdfborder=0 0 0,  %removes outlines around hyper links in online display
%  ]{hyperref}
\usepackage{subfigure}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{verbatim}
% \usepackage[margin=1in]{geometry}
\usepackage{float}
\usepackage{stackrel}
\usepackage{xr}
% \usepackage{microtype}
\usepackage{fancyhdr}
\usepackage{multirow}
\setlength{\headheight}{15.2pt}

\input{defs.tex}


% Approximately 1" margins, more space on binding side
%\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar]{geometry}
%for general printing (not binding)
\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}

% Provides a draft mark at the top of the document. 
% \draftstamp{\today}{DRAFT}


\begin{document} 
\frontmatter

%initialize page style, so contents come out right (see bot) -mjz
\pagestyle{fancyplain}

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Verifying Higher-Order Imperative Programs with Higher-Order Separation Logic}}
\author{Neelakantan R. Krishnaswami}
\date{1 June 2012}
\Year{2012}
\trnumber{CMU-CS-12-127}

\committee{
Jonathan Aldrich, Co-chair \\
John C. Reynolds, Co-chair \\
Robert Harper \\
Lars Birkedal, IT University of Copenhagen
}

\support{This research was partially supported by National Science Foundation Grant
CCF-0916808.}
\disclaimer{The views and conclusions presented in this document are those of the author
and do not reflect the official opinions of the NSF or the U.S. government.}

% copyright notice generated automatically from Year and author.
% permission added if \permission{} given.

\keywords{Separation Logic, Design Patterns, Verification, Specification, Type Theory,
          Denotational Semantics, Program Logic, Program Correctness}

\maketitle

\begin{dedication}
I would like to thank my wife, Rachel, for her love, patience, and
logistical support. 

\ \\

My parents encouraged me to study computers.

I am not at all sure this was what they had in mind. 
\end{dedication}

% \pagestyle{plain} % for toc, was empty
 

\pagestyle{fancyplain}
\renewcommand{\chaptermark}[1]{\markboth{#1}{}}
 
% \lhead{\fancyplain{}{\thepage} --- \texttt{DRAFT 2.167}}
% \lhead{\fancyplain{}{\thepage} --- \texttt{\inp}
\lhead{\fancyplain{}{\thepage}}
\chead{}
\rhead{\fancyplain{}{\textit{\leftmark}}}
\lfoot{}
\cfoot{}
\rfoot{}



%% Obviously, it's probably a good idea to break the various sections of your thesis
%% into different files and input them into this file...

\begin{abstract}
In this thesis I show is that it is possible to give modular
correctness proofs of interesting higher-order imperative programs
using higher-order separation logic.

To do this, I develop a model higher-order imperative programming
language, and develop a program logic for it. I demonstrate the power
of my program logic by verifying a series of examples. This includes
both realistic patterns of higher-order imperative programming such as
the subject-observer pattern, as well as examples demonstrating the
use of higher-order logic to reason modularly about highly aliased
data structures such as the union-find disjoint set algorithm.
\end{abstract}

\begin{acknowledgments}
I would like to begin by thanking my advisors for giving me perhaps
more rope than was advisable. John and Jonathan were both very
tolerant of my sudden enthusiasms, my half-baked ideas, and general
tendency to wander off in peculiar directions. 

Lars was an invaluable source of support and advice, and after every
trip to Denmark I returned significantly more knowledgeable than
before I left. I would like to thank Bob for an act he may not even
recall any more: as a new graduate student, he informed me that I
would be coming to ConCert group meetings, despite technically not
even being part of the ConCert project.
 
As a consequence, I have the need and the pleasure to thank the
whiteboard gang for their willingness to talk. Noam Zeilberger and
Jason Reed were particularly inspiring examples of people following
their mathematical muse, and William Lovas, Daniel Licata, Kevin
Watkins, Joshua Dunfield, Carsten Varming, and Rob Simmons were
reliable sources of type-theoretic and semantic inspiration.
Whenever I needed a different perpsective, the Plaid group was
there. I'd like to thank Donna Malayeri, Kevin Bierhof, and Nels
Beckman in particular.

Hongseok Yang, Peter O'Hearn, Philippa Gardner, and Matthew Parkinson
were all encouraging above and beyond the call of duty to a random
student from thousands of miles away. 

\end{acknowledgments}


\tableofcontents
\listoffigures
% \listoftables

\mainmatter

%% Double space document for easy review:
%\renewcommand{\baselinestretch}{1.66}\normalsize

% The other requirements Catherine has:
%
%  - avoid large margins.  She wants the thesis to use fewer pages, 
%    especially if it requires colour printing.
%
%  - The thesis should be formatted for double-sided printing.  This
%    means that all chapters, acknowledgements, table of contents, etc.
%    should start on odd numbered (right facing) pages.
%
%  - You need to use the department standard tech report title page.  I
%    have tried to ensure that the title page here conforms to this
%    standard.
%
%  - Use a nice serif font, such as Times Roman.  Sans serif looks bad.
%
% Other than that, just make it look good...
% bib

\begin{sloppypar}
\input{introduction}
\input{type-structure}
\input{logic-semantics}
\input{design-patterns}
\input{union-find}
% \input{interactive-programs}
\end{sloppypar}
%\appendix
%\include{appendix}

\backmatter

\input{references}
% 
% %\renewcommand{\baselinestretch}{1.0}\normalsize
% 
% % By default \bibsection is \chapter*, but we really want this to show
% % up in the table of contents and pdf bookmarks.
% \renewcommand{\bibsection}{\chapter{\bibname}}
% %\newcommand{\bibpreamble}{This text goes between the ``Bibliography''
% %  header and the actual list of references}
% \bibliographystyle{plainnat}
% \bibliography{register} %your bib file

\end{document}
